Fractional Types

### 2013-02-02 frac.tex
intro

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
  In previous work, we developed a \emph{first-order},
  information-preserving, and reversible programming language {{langRev}}
  founded on type isomorphisms. Being first-order limits the expressiveness
  of the language: it is not possible, for example, to abstract program
  fragments into a higher-level combinator. More importantly, however, the
  limitation to first-order types means that the class of isomorphisms
  between two types {{b1}} and {{b2}} cannot be itself represented as a type
  and hence that the natural notion of equivalence between two {{langRev}}
  computations is no longer based on isomorphisms. In this paper, we
  introduce a higher-order extension of {{langRev}} based on the novel
  concept of \emph{fractional types}. Intuitively, a value of a fractional
  type represents \emph{negative} information: such fractional values can be
  freely propagated and transformed but must ultimately --- in a complete
  program --- be offset by the corresponding amount of positive information.
\end{abstract}

@@

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
  In previous work, we developed a \emph{first-order},
  information-preserving, and reversible programming language {{langRev}}
  founded on type isomorphisms. Being restricted to first-order types limits
  the expressiveness of the language: it is not possible, for example, to
  abstract common program fragments into a higher-level combinator. In this
  paper, we introduce a higher-order extension of {{langRev}} based on the
  novel concept of \emph{fractional types} {{1/b}}. Intuitively, a value of a
  fractional type {{1/v}} represents \emph{negative} information. A
  higher-order function is modeled by a pair {{(1/v1,v2)}} with {{1/v1}}
  representing the needed argument and {{v2}} representing the
  result. Fractional values can be freely propagated and transformed but must
  ultimately --- in a complete program --- be offset by the corresponding
  amount of positive information.
\end{abstract}

@@

We are witnessing a convergence of ideas from several distinct research
communities (physics, mathematics, and computer science) towards replacing
\emph{equalities} by \emph{isomorphisms}. The combined programme originating
from these research areas has sparked a huge amount of research that unveiled
new and surprising connections between geometry, algebra, logic, and
computation.

In the physics community, Landauer, Feynman, and others viewed the laws of
physics as fundamentally related to computation. In more detail, the great
majority of the laws of physics are formulated as equalities between
different physical observables. For instance, Newton's second law of
classical mechanics equates the force acting on a system to its rate of
change of momentum. The insight of Landauer, Feynman, etc. is that
\emph{different} physical observables should not be related by an
\emph{equality} but more appropriately by an \emph{isomorphism} that
witnesses, explains, and simulates the process of transforming one observable
to the other.

In the mathematics and logic community, Martin-L\"of developed an extension
of the simply typed $\lambda$-calculus originally intended to provide a
rigorous framework for constructive mathematics. It is now natural to extend
this type theory with a new \emph{identity type} representing the proposition
that two terms are ``equal.'' Specifically, given two terms $a$ and $b$ of
the same type $A$, one forms the type $\texttt{Id}_A(a,b)$ representing the
proposition that $a$ and $b$ are equal: in other words, a term of type
$\texttt{Id}_A(a,b)$ witnesses, explains, and simulates the process of
transforming $a$ to $b$ and vice-versa.

In the computer science community, the theory and practice of type
isomorphisms is well-established. Originally, such type isomorphisms were
motivated the pragmatic concern of searching large libraries of functions by
providing one of the many possible isomorphic types for the desired
function. More recently, type isomorphisms have taken a more central role as
\emph{the} fundamental computational mechanism from which more conventional,
i.e., irreversible computation, is derived~\cite{infeffects}. In more detail,
in our own previous work, we started with the notion of type isomorphism as
the ``right'' starting point for a theory of computation. The central
technical contribution was the development of a family of programming
languages, {{langRev}} with various superscripts, in which every computation
is an isomorphism. Other major results are that all such computations
preserve the information-theoretic entropy, that the language is logically
reversible, and that it has a model in symmetric monoidal categories (which
are the backbone of models for linear logic and quantum computation). We have
also invested a significant amount of time in using {{langRev}} as an actual
programming language and established several idioms for programming large
classes of interesting (recursive) programs in {{langRev}}.

A major open problem remains, however. This concerns a higher-order extension
of {{langRev}}. This extension is of fundamental importance in all the
originating research areas. In physics, it allows the process itself or the
observer to be treated as ``data'' that can be transformed by higher-level
processes or observed by meta-level observers. In mathematics and logic, it
allows the equivalence between different proofs of type $\texttt{Id}_A(a,b)$
to be itself expressed as an isomorphism as a higher type
$\texttt{Id}_{\texttt{Id}_A(a,b)}(p,q)$. Finally, in computer science,
higher-order types allow code to abstract over other code fragments as well
as the manipulation of code as data and data as code which is central to many
algorithms and application domains.

We address this higher-order extension in our paper using a simple novel type
construction, that of a fractional type. The plan of the paper is therefore
as follows: \todo{roadmap}
